A Triangle Mesh Approach to Painting

### 2014-04-20 template.tex
Edits to sections 5, 6, and 7.

\documentclass[review]{acmsiggraph}
\usepackage{subcaption}
\usepackage{color}

@@

\begin{CRcatlist}
  %\CRcat{I.3.3}{Computer Graphics}{Three-Dimensional Graphics and Realism}{Display Algorithms}
  %\CRcat{I.3.7}{Computer Graphics}{Three-Dimensional Graphics and Realism}{Radiosity};
\end{CRcatlist}

@@

\subsection{Converting Strokes to Contours}
A set of contours define the outline of a stroke. For simply connected strokes, those without holes, 
one contour describes the whole stroke. However, strokes with holes, such as a circle or 
figure-eight, will have more than one contour.
Converting a user's mouse motion into a set of contours requires two steps.
First, the user's mouse movements are captured by a set of polygons which can be
rendered to yield a rasterized version of the stroke.
Second, to obtain the contours the stroke is rasterized in black and marching squares
is run to extract iso-contours of 50\% gray. This returns a very dense set of points 
which we reduce through a pruning process to
get a sparse but accurate representation of the contour.

@@

We utilized Jonathan Shewchuk's Triangle library to perform Delaunay triangulations
on the given set of points and boundaries. The library outputs a mesh that
approximately represents the drawn stroke.

@@

\subsection{Soft Strokes}
The above procedure converts a hard brush into a triangle mesh. Yet, a somewhat more interesting
case is converting a soft brush with linear gradient into a triangle mesh. A soft stroke has two
components, an inner hard stroke and an outer gradient stroke. Since the triangles in the inner and
outer parts of the stroke must be colored differently, none of the triangles may cross the boundary
between the hard and soft regions.

@@

Again, using marching squares these contours are extracted and then pruned. The boundaries and points
derived from the contours are fed into Triangle and a mesh is returned. This mesh represents the soft
stroke where no triangle crosses the boundary between the soft and hard parts
of the stroke. Points on the contour of the inner stroke will have color arcs with the stroke color
while the points on the contour of the outer stroke will have transparent color arcs. This gives
all triangles connecting the inner and outer strokes a linear gradient.

@@

Every canvas begins with a number of points, along the edge of the screen. These
points have boundaries connecting them. These points are originally part of a number of triangles which
divide the canvas.

In the simplest approach the points and boundaries from the new stroke are added
to the list of points and boundaries already stored. All of this data can be fed into Triangle
to produce a new mesh. The new mesh contains all of the points of the old and new strokes. Since
all of the boundaries are stored and fed into Triangle the stroke edges are preserved. This means that
no triangle can cross the boundary of a stroke. Without this constraint it would be impossible to color the
triangles since they would straddle color arcs. Once a new mesh is formed the color arcs of the 
new and old points must be computed.

@@

\subsection{Determining Colors}
The color arcs for new points depend on two values, the color of the stroke and the color
of the canvas at that point. The points on a hard stroke will have two color arcs. One
arc describes the inside of the stroke. The color for this arc is the color of the stroke
composited over the color of the canvas at that point. The other arc describes the outside
of the stroke. The color for this arc is just the color of the canvas at that point. The
points on a soft stroke will have one color arc. The color arc for points on the inner
contour is the color of the stroke composited over the color of the canvas at that point.
The color arc for points on the outer contour is the color of canvas at that point.
This gives any triangle connecting inside and outside contour points a linear gradient
from the stroke color composited over the canvas color to the canvas color.

@@

When Triangle is asked to triangulate a set of points with overlapping boundaries it must
insert a point at the intersection, since without it at least one triangle would have to
cross a boundary. These points are introduced with no color information, but the final colors for 
all of the other points in the mesh have already been determined.

An intersection point lies on two boundaries and therefore has four boundary edges emanating from it. 
To determine the colors of an intersection points requires knowing the colors associated with
points on the boundaries. For both boundaries travel both directions to the first point with color data. Note this may not
be the first point on each side since this intersection point may be adjacent to other intersection
points without color data. For both boundaries the colors in the color arcs are linearly interpolated. 
This yields four average colors, two from the new stroke boundary and two from the old stroke boundary.
The colors from the new stroke are composited over the colors from the old stroke to yield four new
colors, one for each region defined by the boundary intersections.

@@

\subsection{Finding Modified Points}
The above procedure has two important drawbacks. First, it processes many points that don't need
consideration which takes a considerable amount of processing time. Second, since it acts on the whole
canvas, local changes can have global effects. To avoid this problem the algorithm determines
which triangles can remain
and which triangles need to be included in the re-triangulation. To accomplish this a static grid
is generated and all of the cells which contain a triangle from the new stroke are marked. 
The set of triangles from the old triangulation that intersect these grid squares are found.
The set of edges from these triangles that do not intersect these grid cells form a ring
around the area of our new triangles. The edges in this ring become constraints
in the triangulation. All of the points and constraint edges from the old triangulation are
included as well as the points and constraint edges from the new triangulation. Triangulating
these points and edges gives the new geometry for the modified area and has no effect on any
other part of the canvas. Therefore, the old triangles from the rest of the canvas can be reused. 
The new triangulation combined with the old triangles give a new triangle mesh representation of the entire canvas.

\begin{figure}
    \centering
        \includegraphics[height=1.5in]{images/findingmodifiedpoints}
    \caption{Place Holder.}
\end{figure}

\section{Discussion and Results}
Here we first consider how the complexity of the algorithm grows both in time and geometry.
The results from this analysis and our experience with TrianglePainter have lead to observations in three main areas.

\subsection{Complexity}
To find the growth rates of the algorithm we performed two experiments. 
First, the program drew fifty random strokes without zooming.
Next, the program drew fifty random strokes with random zooming.
Geometry grows fastest when many strokes are composited on one another. When not zooming strokes
must cover the same area of canvas over and over. This
results in more geometry and therefore takes longer to render. The zooming case explores different
areas of the canvas and therefore results in less overlapping geometry. This situation should result
in less complex geometry and therefore take less time.

\begin{figure}
  \centering
  \begin{subfigure}[b]{0.5\columnwidth}
    \centering
    \includegraphics[width=\columnwidth]{graphs/numtriangles}
    \label{fig:numtriangles}
  \end{subfigure}%
  \begin{subfigure}[b]{0.5\columnwidth}
    \centering
    \includegraphics[width=\columnwidth]{graphs/cumulativetime}
    \label{fig:timing}
  \end{subfigure}%
  \vspace{-0.4cm}
  \caption{For paintings of 50 random strokes, the left graph shows the number of triangles in the canvas as of stroke $x$, and the right graph shows the cumulative render time of the painting. The blue points are the non-zooming case, while the red points are the zooming case.}
\end{figure}

As seen in figure~\ref{fig:numtriangles} the zooming case grows linearly while the non-zooming
appears to grow quadratically. This is expected since every new stroke has the chance to overlap
all of the old strokes. This means each new stroke does not just add a constant amount of geometry.
Due to intersections with previous strokes intersection points are added. This results in a linear
increase in the amount of geometry. This results
in a quadratic growth rate. However, in the zooming case each stroke rarely overlaps and therefore
intersection points are rarely added. Instead the geometry increases by a constant which leads to
a linear growth rate.

Figure~\ref{fig:timing} gives a similar result for the cumulative time strokes take to render. Both
the zooming and non-zooming cases grow non-linearly over time. However, the non-zooming case takes
a lot longer for the reasons described. It's not unexpected that even the zooming case grows non-linearly
since even though the geometries involved aren't getting very complicated, the geometry of the entire
canvas is increasing linearly and must be considered even if the entire canvas is not retriangulated.

All this suggests that a simplification is of great importance.
The case that it is most likely to help is the non-zooming case with complex geometry. When lots of strokes
overlap there is usually an extra geometry that does not add much detail to the drawing. By removing
this extra geometry the algorithm should achieve a steady growth rate.

\subsection{Stability}
Something we've worked hard to achieve is program stability. Certain geometries are fatal
for the Triangle library. The current instability in the program is normally caused by
providing the library with such difficult geometries. To improve program stability would
require addressing geometries such as vertices that are too close together. This would
take more processing time and is something we have yet to explore.

\subsection{Performance}
The current implementation uses Python and PyOpenGL. Python's performance is slower than
a lower level implementation in C++ would be. The drawing of the canvas is quick with
no lag. However, the actual process to take a stroke from rasterization to triangle
representation on the canvas is slow. On a blank canvas strokes take on the order of
1 second to process. However, on a canvas with lots of geometry a stroke that covers
that geometry can take several seconds to render. Some of this time is due to poor
implementation of the algorithm. However, there are lots of calculations
to do with the old points affected by the new stroke and the points associated with the
new stroke itself. Since these geometries can get arbitrarily complicated, these
calculations can take an arbitrarily long. This brings us to the final point, scalability.

\subsection{Simplification}
No matter how good the implementation, the fact that paintings can get arbitrarily complicated
means there must be some way to simplify the mesh. Although we have yet to implement
such a simplification algorithm, we have a couple suggestions on how it might be done. In most situations where
a lot of strokes are composited on top of one another, most of the geometry is redundant. 
For example, when an opaque stroke is laid down on top of a complex geometry, that complex
geometry no longer has any useful information since the outline of the opaque stroke
completely defines it. Furthermore, even with translucent or feathered strokes, enough
composited strokes may produce complex geometry that does not add much to the output.
Some of this geometry can be reduced without noticeable changes in the output. Such a
simplification algorithm must be implemented to ensure a painter can continue to paint
without the program slowing to an unusable state.

@@

% - Switch 6.3 and 6.4
\begin{figure*}
    \centering
        \includegraphics[width=0.49\textwidth]{images/facezoom}
        \includegraphics[width=0.49\textwidth]{images/facezoomtriangles}
    \caption{Artist time ~30 minutes. 9030 triangles. \Steve{Can we add number of strokes?}}
\end{figure*}

\begin{figure*}
    \centering
        \includegraphics[width=0.195\textwidth]{images/zoom61}
        \includegraphics[width=0.195\textwidth]{images/zoom62}
        \includegraphics[width=0.195\textwidth]{images/zoom63}
        \includegraphics[width=0.195\textwidth]{images/zoom64}
        \includegraphics[width=0.195\textwidth]{images/zoom65}
        \includegraphics[width=0.195\textwidth]{images/zoom66}
        \includegraphics[width=0.195\textwidth]{images/zoom67}
        \includegraphics[width=0.195\textwidth]{images/zoom68}
        \includegraphics[width=0.195\textwidth]{images/zoom69}
        \includegraphics[width=0.195\textwidth]{images/zoom71}
    \caption{A painting that shows extreme zoom levels.  The final image is at \Steve{zoom level XXX}.}
\end{figure*}

\section{Conclusion}
Vector graphics are useful for representing complex images with limited information
while also enabling easy editing and infinite resolution. Previous work has focused
on converting raster graphics to vector graphics, unique vector representations, and multi-scale
raster graphics. However, until now no system
has enabled artists to create an entire painting in vector graphics using standard brush techniques.

TrianglePainter enables a user to paint like they would in a raster graphics program 
to create vector graphics. The core of TrianglePainter is a triangle mesh representation of the image.
We have created an algorithm to convert a user's mouse motions into a triangle mesh, and then composite
that mesh on the preexisting mesh containing the previous strokes.

The current implementation allows artists to create minimal works. Future work should focus on
implementing a simplification algorithm to allow the complexity of the drawings to grow. Other
improvements might focus on creating interesting transforms, or other operations that utilize
the underlying triangle mesh structure to make the artists' job easier.

>>>>

\documentclass[review]{acmsiggraph}

\usepackage{subcaption}
\usepackage{color}

% Reduce the space between subfigures and subfigure captions.
\captionsetup[subfigure]{aboveskip=6pt}
\captionsetup[subfigure]{belowskip=0pt}

@@

\begin{CRcatlist}
  \CRcat{I.3.4}{Computer Graphics}{Graphics Utilities}{Paint systems}
  \CRcat{I.3.6}{Computer Graphics}{Methodology and Techniques}{Graphics data structures and data types}
\end{CRcatlist}

@@

\subsection{Converting Strokes to Contours}
A set of contours define the outline of a stroke. For simply connected strokes, those without holes, 
one contour describes the whole stroke. However, strokes with holes, such as a circle or 
figure-eight, will have more than one contour.
Converting a user's mouse motion into a set of contours requires two steps.
First, the user's mouse movements are captured by a set of polygons which can be
rendered to yield a rasterized version of the stroke.
Second, to obtain the contours the stroke is rasterized in black and marching squares~\cite{XXX}
is used to extract iso-contours of 50\% gray. This returns a very dense set of points 
which we reduce through a pruning process to
get a sparse but accurate representation of the contour.

@@

We utilized Jonathan Shewchuk's Triangle library~\cite{XXX} to perform constrained Delaunay triangulations
on the given set of points and boundaries. The library outputs a mesh that
approximately represents the drawn stroke.

@@

\subsection{Soft Strokes}
The above procedure converts a hard brush stroke into a triangle mesh. A more interesting
case is converting a soft brush stroke with linear gradient into a triangle mesh. A soft stroke has two
components, an inner hard stroke and an outer gradient stroke. Since the triangles in the inner and
outer parts of the stroke must be colored differently, none of the triangles may cross the boundary
between the hard and soft regions.

@@

Again, using marching squares these contours are extracted and then pruned. The boundaries and points
derived from the contours are given to Triangle and a mesh is returned. This mesh represents the soft
stroke where no triangle crosses the boundary between the soft and hard regions
of the stroke. Points on the contour of the inner stroke will have color arcs with the stroke color,
while the points on the contour of the outer stroke will have transparent color arcs. This gives
all triangles connecting the inner and outer strokes a linear gradient (see Figure~\ref{XXX}).

@@

Every canvas begins with an initial set of points distributed along the edges. These
points have boundaries connecting them. These points are originally part of a number of triangles which
divide the canvas.

In the simplest approach, the points and boundaries from the new stroke are added
to the list of points and boundaries already stored on the canvas. All of this data is given to Triangle
to produce a new mesh. The new mesh contains all of the points of the old and new strokes. Since
all of the boundaries are stored and given to Triangle, the stroke edges are preserved. This means that
no triangle can cross the boundary of a stroke. Without this constraint, it would be impossible to correctly color the
triangles since they would straddle color arcs. Once a new mesh is formed the color arcs of the 
new and old points must be computed.

@@

\subsection{Determining Colors}
All color compositing is done with regular RGB alpha blending.  The color arcs for new points depend on two values, the color of the stroke and the color
of the canvas at that point. The points on a hard stroke will have two color arcs. One
arc describes the inside of the stroke. The color for this arc is the color of the stroke
composited over the color of the canvas at that point. The other arc describes the outside
of the stroke. The color for this arc is just the color of the canvas at that point. The
points on a soft stroke will each have one color arc. The color arc for points on the inner
contour is the color of the stroke composited over the color of the canvas at that point.
The color arc for points on the outer contour is the color of canvas at that point (because the outer edge of the soft stroke is completely transparent).
This gives any triangle connecting inside and outside contour points a linear gradient
from the stroke color composited over the canvas color to the canvas color.

@@

When Triangle is asked to triangulate a set of points with intersecting boundary edges, it must
insert a point at the intersection, since without it at least one triangle would have to
cross a boundary. These points are introduced with no color information, but the final colors for 
all of the other points in the mesh have already been determined.

An intersection point lies on two boundaries and therefore has four boundary edges emanating from it. 
To determine the colors of an intersection points requires knowing the colors associated with
other points on the boundaries. For both boundaries, travel both directions to the first point with color data. Note this may not
be the first point on each side since this intersection point may be adjacent to other intersection
points also without color data. For both boundaries, the colors in the color arcs are linearly interpolated. 
This yields four average colors, two from the new stroke boundary and two from the old stroke boundary.
The colors from the new stroke are composited over the colors from the old stroke to yield four new
colors, one for each region defined by the boundary intersections.

@@

\subsection{Finding Modified Points}
The above procedure has two important drawbacks. First, it processes many points that do not need
consideration which takes a considerable amount of processing time. Second, since it acts on the whole
canvas, local changes can have global effects. To avoid this problem the algorithm determines
which triangles can remain
and which triangles need to be included in the re-triangulation (see Figure~\ref{fig:modified}). To accomplish this a static grid
is generated and all of the cells which contain a triangle from the new stroke are marked. 
The set of triangles from the old triangulation that intersect these grid squares are found.
The set of edges from these triangles that do not intersect these grid cells form a ring
around the area of our new triangles. The edges in this ring become constraints
in the triangulation. All of the points and constraint edges from the old triangulation are
included as well as the points and constraint edges from the new triangulation. Triangulating
these points and edges gives the new geometry for the modified area and has no effect on any
other part of the canvas. Therefore, the old triangles from the rest of the canvas can be reused. 
The new triangulation combined with the old triangles give a new triangle mesh representation of the entire canvas.

\begin{figure}
    \centering
        \includegraphics[height=1.5in]{images/findingmodifiedpoints}
    \caption{Place Holder.}
    \label{fig:modified}
\end{figure}

\section{Discussion and Results}
Paintings made with TrianglePainter can be seen in Figures~\ref{fig:face} and \ref{fig:boxes}.  Here we first consider how the complexity of the algorithm grows both in time and geometry.
The results from this analysis and our experience with TrianglePainter have lead to a number of observations.

\subsection{Complexity}
To find the growth rates of the algorithm we performed two experiments. 
First, the program drew fifty random strokes without zooming.
Next, the program drew fifty random strokes with random zoom level for each stroke.
Geometry grows fastest when many strokes are composited on one another. When not zooming, strokes
must cover the same area of canvas over and over. This
results in more geometry and therefore takes longer to render. The zooming case explores different
areas of the canvas and therefore results in less overlapping geometry. This situation should result
in less complex geometry and therefore take less time.

\begin{figure}
  \centering
  \begin{subfigure}[b]{0.5\columnwidth}
    \centering
    \includegraphics[width=\columnwidth]{graphs/numtriangles}
    \caption{Number of triangles}
    \label{fig:numtriangles}
  \end{subfigure}%
  \begin{subfigure}[b]{0.5\columnwidth}
    \centering
    \includegraphics[width=\columnwidth]{graphs/cumulativetime}
    \caption{Rendering time}
    \label{fig:timing}
  \end{subfigure}%
  \caption{For paintings of 50 random strokes, the left graph shows the number of triangles in the canvas as of stroke $x$, and the right graph shows the cumulative render time of the painting. The blue points are the non-zooming case, while the red points are the zooming case.}
\end{figure}

Figure~\ref{fig:numtriangles} shows the zooming case grows linearly while the non-zooming case
appears to grow quadratically. This is expected since every new stroke has the chance to overlap
all of the old strokes. This means each new stroke does not just add a constant amount of geometry.
Due to intersections with previous strokes intersection points are added. This results in a linear
increase in the amount of geometry. This results
in a quadratic growth rate. However, in the zooming case each stroke rarely overlaps and therefore
intersection points are rarely added. Instead the geometry increases by a constant which leads to
a linear growth rate.

Figure~\ref{fig:timing} gives a similar result for the cumulative time strokes take to render. Both
the zooming and non-zooming cases grow non-linearly over time. However, the non-zooming case takes
longer for the reasons described. It is not unexpected that even the zooming case grows non-linearly
since even though the geometries involved are not becoming very complicated, the geometry of the entire
canvas is increasing linearly and must be considered even if the entire canvas is not re-triangulated.

This suggests that canvas simplification is of great importance.
The case that it is most likely to be helped is the non-zooming case with complex geometry. When lots of strokes
overlap there is usually extra geometry that does not add much detail to the drawing. By removing
this extra geometry the algorithm should achieve a steady growth rate.

\subsection{Stability}
While the Triangle library is impressively robust, there are still certain pathological geometries that can cause it to fail.  Specifically, long skinny triangles and vertices that are too close together seem to trigger triangulation failures.  Detecting these situations and fixing them, or even better modifying the geometry construction to avoid them, would solve this problem.  For example, identifying if a new vertex is too close to an existing one, and if so, merging them.  Another example is to subdivide long edges to reduce the incidence of skinny triangles.  The impact these solutions would have on processing time is something that needs to be explored.

\subsection{Performance}
The current implementation uses Python and PyOpenGL. Python's performance is slower than
a lower level implementation in C++ would be. The drawing of the canvas is quick with
no lag. However, the actual process to take a stroke from rasterization to triangle
representation on the canvas can be slow. On a blank canvas, strokes take less than
one second to process. However, on a canvas with lots of geometry a stroke that overlaps
that geometry can take several seconds to render. Some of this could be alleviated by a carefully optimized implementation. However, there are many calculations
necessary for the old points affected by the new stroke and the points associated with the
new stroke itself. Since these geometries can become arbitrarily complicated (in pathological cases), these
calculations can take an arbitrarily long. This brings us to the final point, scalability.

\subsection{Simplification}
No matter how good the implementation, the fact that paintings can become arbitrarily complicated as more and more strokes are added
means there must be some way to simplify the mesh. In most situations where
many strokes overlap one another, most of the geometry is redundant. For example, when an opaque stroke is laid down on top of a complex geometry, that complex
geometry no longer has any useful information since the outline of the opaque stroke
completely defines it (all vertices are the same color).  Furthermore, even with translucent or feathered strokes, a large enough number of 
composited strokes may produce complex geometry that does not add much to the visual appearance of the canvas. Some of this geometry can be reduced without noticeable changes.  To accomplish this, our proposed simplification algorithm would identify a vertex that can be removed via an edge collapse by measuring the change the collapse would cause in the image.  The change is computed as the magnitude of the color shift and the translation of a boundary.  If these deltas are below a threshold, then the collapse can proceed.  This process could be done when a stroke is added to the canvas, or continually in a background thread to avoid increasing apparent latency.  Though we have yet to implement this simplification algorithm, it is important to ensure a painter can continue iterate on a painting without the program becoming too slow.

@@

% - Switch 6.3 and 6.4
\begin{figure*}
    \centering
        \includegraphics[width=0.49\textwidth]{images/facezoom}
        \includegraphics[width=0.49\textwidth]{images/facezoomtriangles}
    \caption{Artist time ~30 minutes. 9030 triangles. \Steve{Can we add number of strokes?}}
    \label{fig:face}
\end{figure*}

\begin{figure*}
    \centering
        \includegraphics[width=0.195\textwidth]{images/zoom61}
        \includegraphics[width=0.195\textwidth]{images/zoom62}
        \includegraphics[width=0.195\textwidth]{images/zoom63}
        \includegraphics[width=0.195\textwidth]{images/zoom64}
        \includegraphics[width=0.195\textwidth]{images/zoom65}
        \includegraphics[width=0.195\textwidth]{images/zoom66}
        \includegraphics[width=0.195\textwidth]{images/zoom67}
        \includegraphics[width=0.195\textwidth]{images/zoom68}
        \includegraphics[width=0.195\textwidth]{images/zoom69}
        \includegraphics[width=0.195\textwidth]{images/zoom71}
    \caption{A painting that shows extreme zoom levels.  The final image is at \Steve{zoom level XXX}.}
    \label{fig:boxes}
\end{figure*}

\section{Conclusion}
TrianglePainter enables a user to paint like they would in a raster graphics program 
to create vector graphics. The core of TrianglePainter is a triangle mesh representation of the image.
We have created an algorithm to convert a user's mouse motions into a triangle mesh, and then composite
that mesh on the preexisting mesh containing the previous strokes.

The current implementation provides a baseline painting experience. Future work should focus on
implementing a simplification algorithm to allow the complexity of the drawings to grow. Other
improvements might focus on creating interesting manipulation tools, or other operations that utilize
the underlying triangle mesh structure to make the artists' job easier.
